The Spatial and Temporal Domains of Modern Ecology

### 2017-12-12 paper/main.tex
First round of edits for paper main text through Table 1 in intro.

% Place your abstract within the special {sciabstract} environment.

%\begin{sciabstract}
\section*{Abstract}
%Ecologists have long known that the observational scales play a critical role in shaping our understanding of ecosystems, and that observations spanning multiple scales of space and time are needed to effectively characterize ecological phenomena. 
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 studies published between 2004-2014. We found that the scale domains of observations are fairly narrow, and are collected primarily with conventional field techniques. In the spatial domain, most observations have resolutions of $\leq$1 m2 and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and were made over relatively short durations ($\leq$1 year). Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, raising concerns about observational representativeness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that it is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

%as well as \emph{actual} extent (the summed resolution of spatial replicates) and \emph{actual} duration (integrated sample duration of temporal replicates). Our results suggest that the scale domains of modern ecological observations remain fairly narrow, and are collected primarily with conventional field techniques. A comparison with results from prior earlier, scale-focused meta-analyses revealed that observational durations and resolutions are largely unchanged since the 1980s, but intervals have become finer and extents larger, the latter driven by steadily increasing use of remote sensing.
%Most observations had resolutions of $\leq$1 m$^2$ (67\%) and extents $\leq$10,000 ha (53\%). Temporally, most observations were either unreplicated (37\%) or had intervals $\geq$1 month (26\%), while 64\% of observations had durations of $\leq$1 year. Juxtaposing dimensions revealed several notable ``holes'' (domains with few to no observations), particularly in the realms of high frequency (sub-daily to daily intervals) observations with fine to moderate resolutions ($\geq$1 m$^2$ up to 100 ha) and $\geq$decadal durations. These holes relate to tradeoffs between observational frequency and resolution, as well as duration, that are characteristic of field-based data collection, which was used for 80\% of observations. An analysis of values reported in the most comparable prior scale meta-analyses revealed that observational durations and resolutions are largely unchanged since the 1980s, but intervals have become finer and extents larger. Our data also showed an internal trend of increasing observational extent, coinciding with a 1.3\% per year increase in remote sensing-based observations. Despite this trend, the \emph{actual} extent of 81\% of observations was 4-8 orders of magnitude smaller than extent. Similarly, the \emph{actual} duration of 67\% of observations was 3-9 orders of magnitude shorter than duration. These large discrepancies raise questions about the representativeness of observations, which may in part be addressed by examining the degree to which auto-correlation processes close the gap between actual and represented scales of observation. Adding to this concern is the fact that most observations did not clearly report observational scales, suggesting that scale remains a minor consideration throughout much of ecology.}

%Observational scales play a critical role in shaping our understanding of ecological pattern and process. To gain insight into the spatial and temporal domains being addressed by modern ecological science, we conducted a meta-analysis of papers published in 30 ecological journals between 2004 and 2014. We estimated two spatial dimensions (resolution and extent) and two temporal dimensions (interval and duration) of 371 ecological observations reported within 140 papers (out of 346 randomly selected titles). Our analysis revealed that ecology largely remains a field-based discipline that primarily collects observations within narrow spatial and temporal domains. Ecologists still make limited use of methods for observing larger-scale patterns and processes (e.g. remote sensing, used for 6\% of observations), but make greater use of those which permit high frequency or temporally continuous \emph{in situ} measurements (12\% of observations). Resolution, extent, and duration were not clearly reported for the majority of observations, which, together with the narrow spatio-temporal ranges of observed scales, suggests a limited ability to generalize the findings of many studies.

%  An example of the style is the special\texttt{\{sciabstract\}} environment used to set up the abstract you see here.
%\end{sciabstract}

@@

%One of the best known and most comprehensive assessments of scale in ecological research ident

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses of ecological research scales were published nearly 30 years ago \cite{tilman_ecological_1989,kareiva_spatial_1988}, it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains of empirical observations that were reported within recently (2004-2014) published ecological studies (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}). Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

%This question may be best addressed by evaluating the scales of empirical observations. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see SI for full definitions). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches. Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation. We also assessed observational scales within two additional dimensions, \emph{actual} extent (the integrated area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from those they are intended to represent.

%For example, consider a hypothetical study reporting estimates of herbaceous plant cover in a forest understory, in which the authors reported that they placed 10 400 m$^2$ plots throughout the forest, and within each plot randomly placed 10 1 m$^2$ quadrats, within which they estimated percent herbaceous plant cover. In this example, we consider 1 m$^2$ to be the resolution.

\begin{table}[htp]
\caption{The scale dimensions of ecological observations assessed in this meta-analysis.}
\begin{center}
\begin{tabular}{llll}
\hline
\multicolumn{2}{c}{\textbf{Component}} & \textbf{Units} & \textbf{Description} \\\hline\hline
\multirow{3}*{Spatial} & Resolution & m$^2$ & Area of an individual spatial replicate (plot) \\%\cmidrule(l){2-3}
   & Extent & ha & Area  encompassed by all spatial replicates \\
   & Actual extent & ha & Summed area of all spatial replicates\\\hline
\multirow{3}*{Temporal} & Interval & days & Time elapsed between successive temporal replicates \\
   & Duration & days & Time elapsed between first and last temporal replicates\\
   & Actual duration & days & Summed observational time of all temporal replicates\\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%

>>>>

\section*{Abstract}
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 observational studies published between 2004-2014. We found that the scale domains of observations are fairly narrow, and are collected primarily with conventional field techniques. In the spatial domain, most observations have resolutions of $\leq$1 m2 and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and were made over relatively short durations ($\leq$1 year). Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, revealing unobserved portions of space and time not truly captured by replicates and raising concerns about observational comprehensiveness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that it is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

@@

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses of ecological research scales were published nearly 30 years ago \cite{tilman_ecological_1989,kareiva_spatial_1988}, it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}) of empirical observations (defined here as ecological observations collected under un-controlled or non-manipulated conditions) that were reported within recently (2004-2014) published ecological studies. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see SI for full definitions). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches. Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation. We also assessed observational scales within two additional dimensions, \emph{actual} extent (the summed area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from the scales that the observations are intended to cover. This difference may contain important information about how effectively ecological observations characterize ecological phenomena. First, the odds that observations do not adequately represent the phenomenon of interest may increase (due to inappropriate sampling) as the gap between the actual and intended observational scales increases. Second (and related), a larger gap may increase the likelihood of collecting censored data (the values collected by observations are not fully known), as the frequent complexity, non-linearity, and non-randomness of ecosystems is such that phenomena may resolve themselves in the space or time between observations.

\begin{table}[htp]
\caption{The scale dimensions of ecological observations assessed in this meta-analysis.}
\begin{center}
\begin{tabular}{llll}
\hline
\multicolumn{2}{c}{\textbf{Component}} & \textbf{Units} & \textbf{Description} \\\hline\hline
\multirow{3}*{Spatial} & Resolution & m$^2$ & Area of an individual spatial replicate (e.g. plot) \\%\cmidrule(l){2-3}
   & Extent & ha & Area  encompassed by all spatial replicates \\
   & Actual extent & ha & Summed area of all spatial replicates\\\hline
\multirow{3}*{Temporal} & Interval & days & Time elapsed between successive temporal replicates \\
   & Duration & days & Time elapsed between first and last temporal replicates\\
   & Actual duration & days & Summed observational time of all temporal replicates\\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%
